Diversity and evolution of centromere repeats in the maize genome

### 2014-05-05 centc.tex
Figure names changed
Jose affil added, Addressed major JRI comments in disc

\institute{Paul Bilinski : Kevin Distor : Jose Gutierez-Lopez : Gabriela Mendoza Mendoza : Jeffrey Ross-Ibarra\at
              Department of Plant Sciences, UC Davis\\
              Davis, California, 95616
              \and
              Jeffrey Ross-Ibarra\at
              The Genome Center and Center for Population Biology, University of California\\
              Davis, California, 95616\\
              \email{rossibarra@ucdavis.edu}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Jinghua Shi : R. Kelly Dawe\at
           Department of Plant Biology, University of Georgia\\
	  Athens, Georgia, 30602
           \and
           R. Kelly Dawe\at
           Department of Genetics, University of Georgia\\
           Athens, Georgia, 30602\\
}

@@

Centromere repeats are found in most eukaryotes and play a critical role in kinetochore formation.  Though they exhibit considerable diversity both within and among species, little is understood about the mechanisms that drive centromere repeat evolution.  Here, we use maize as a model to investigate how a complex history involving polyploidy, fractionation, and recent domestication has impacted the diversity of the maize CentC repeat.  We first validate the existence of long tandem arrays of repeats in maize and other taxa in the genus \emph{Zea}.  We find that genetic similarity among CentC copies is highest within these arrays, suggesting that tandem duplications are the primary mechanism for the generation of new copies.  In spite of this, we see little evidence that CentC variants form distinct genetic groups, instead finding that homoplasious mutations have likely homogenized CentC diversity.   Although the two ancestral subgenomes of maize have contributed nearly equal numbers of centromeres, our analysis shows that a vast majority of all CentC repeats derive from only one of the parental genomes.  Finally, by comparing maize with its wild progenitor teosinte, we find that the abundance of CentC, but not the pericentromeric repeat Cent4, has decreased through domestication.

@@

\section*{Introduction}
\label{intro}
In spite of the rapid growth of sequenced plant genomes, plant centromeres remain poorly understood and relatively cryptic, due largely to their highly repetitive content.  Centromere repeats are highly diverse across taxa and their turnover appears to be very rapid \citep{Melters2012}. However, little is known about the genetic mechanisms that produce centromere repeat diversity.  
Domesticated maize (\emph{Zea mays} ssp. \emph{mays}) has a high quality genome assembly \citep{Schnable2009} including complete sequence of two centromeres \citep{Wolfgruber2009}, and the breadth of research into maize centromeres makes it one of the best systems in which to investigate the processes governing centromere repeat evolution.

@@

%We wanted to better understand the diversity present in all CentC copies across the genome.  Looking at our alignment, we found that the sequences had a pairwise percent identify of 65.8\% with only 2.2\% of sites identical, suggesting large sequence diversity.  However, these statistics do not inform us whether the amount of diversity observed would be expected due to cumulative mutation over time.  
In order to better understand patterns of diversity at CentC, we performed simulations to test the likelihood of homoplasious mutations (i.e. independent mutations occurring at the same position in two different CentC repeats). Our simulation (code available at: \url{https://github.com/paulbilinski/CentC_Analyses/tree/master/Diversity_sims}) assumed that CentC has been evolving for  1 million years since the divergence of maize and \emph{Tripsacum} \citep{Ross-Ibarra2009}, a closely related genus whose centromere repeat shares a large amount of homology \citep{Melters2012}.  We assumed a constant copy number, a mutation rate of $3 x 10^{-8}$ \citep{clark2005}, and one generation per year.

@@

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
\includegraphics{Fig1_circos.png}
\caption{CentC repeat location in relation to the maize subgenomes.  The outer ring depicts chromosomal assignment to the two subgenomes, with higher confidence regions are colored with darker colors.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty. The middle ring, shaded in blue, displays the locations of all CentCs across the genome.  The inner ring, shaded in yellow, displays the locations of all CentC clusters greater than 20KB in length}
\label{circos}    
\end{figure}

@@

\begin{figure}[h]
\centering
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Fig3_Heatmap}
\caption{CentC physical location and genetic relatedness for (a) chromosome 2 and (b) chromosome 5.  On the physical map, red lines show locations of numbered CentC clusters and blue blocks show the location of the active kinetichores.  Scale bar is in MB.  Below each physical map is shown a heatmap of genetic relatedness of each CentC to (top row) other copies within its island of tandem repeats delineated by dotted lines and (bottom row) all other copies on the chromosome.  Darker colors indicate higher relatedness.  The total number of CentC in each cluster is shown below the map
}
\label{heatmap}    
\end{figure}

@@

\jri{check tense throughout. i think some is past and some is present. also fill in species names X and Y and add CITEs} 
Our analysis of centromere repeat diversity across the maize genome identifies thousands of copies exhibiting tremendous diversity. 
But while we can cluster the repeats into groups of related sequences, these groups seem to have little relation to current or ancient maize chromosomes (Figure \ref{pcoa}).  
We see no evidence of chromosome specific repeats such as that seen in X and Y \citep{Kawabe2005, Pontes2004}, suggesting the presence of a mechanisms that homogenizes repeats across centromeres on different chromosomes. 
We further show that Cent4, once thought to be a chromosome-specific centromere repeat (CITE), appears to be a poorly characterized tandem repeat or nonautonomous retroelement, but is not associated with the centromere.  
Inter-chromosomal transposition was postulated by \citet{Shi2010} as an explanation for marker genotypes showing evidence of recombination, and such transposition could also explain the similarity of repeats across all chromosomes.

Instead of chromosome-specific groups, we find that virtually all the large arrays of CentC in the maize  genome derive from one of the two ancestral genomes present in modern day \emph{Zea} (Figure \ref{circos}).  
One possible explanation for this result is that centromere inactivation mirrors the differential expression seen between the subgenomes.  
Centromere inactivation could then lead to differential loss of CentC clusters, such that large clusters remain only in the active centromeres.
Large CentC clusters are clearly biased in favor of subgenome 1, which is known to have lower gene loss and higher average expression than maize subgenome 2 \cite{Schnable2011}. 
Consistent with this idea, \citet{ReinhartBartel2002} find evidence of regulatory microRNA’s corresponding to centromeric repeats.   
Of course, an alternative explanation is simply that the subgenome 1 parent had more or larger CentC arrays, leading to an initial unequal representation in the ancient polyploid ancestor of maize.

@@

%!RELATEDNESS
Our sequence comparison of CentCs also enabled us to explore the relationship between genetic and physical distance among repeats.  
Using the well-assembled centromeres on chromosomes 2 and 5, we found spatial autocorrelation of relatedness among repeats, and we observe genome-wide that many CentC’s within an array fall into the same genetic cluster.  
These observations are consistent with the simple idea that most repeats arise due to tandem duplication, and long distance transposition of CentC, while necessary to homogenize repeats across chromosomes \citep{Shi2010}, is relatively uncommon.

@@

\jri{insert CITE, italics, probably refs to tables/figs as appropriate}  Previous cytogenetic work  identified  differences in centromere repeat content  between domesticated maize and its wild relatives Z. mays ssp. parviglumis,), Z. mays ssp. mexicana,, and Z. luxurians (hereafter parviglumis, mexicana, and luxurians) \citep{Albert2010} but was unable to quantify these differences. Our resequencing results show that while there is little difference in the distribution of CentC in tandem arrays, the total abundance of CentC has  decreased  \jri{can we say absolute abundance? we have genome size estimates, right?} during domestication, and we verify this with FISH in two maize-teosinte hybrid individuals (Figure \ref{abundance}). This result contrasts with the apparent increase in transposable element abundance  \citep{Chia2012}, suggesting that changes in CentC are not due to causes common to all repetitive sequence. \jri{ are we sure TEs have increase in absolute number? didn't chia et al. look at RPKM?  wouldn't that increase if overall genome size decreases? if we control for genome size, does TE content show and absolute increase?} The maize genome is smaller than its teosinte counterpart, largely due to differences in the abundance of heterochromatic knobs (Poggio et al 1998).  \citet{Zhang2012} have postulated an adaptive relationship between centromere size and genome size based on an observed correlation between centromere size and genome size across a number of grass species.  This model would propose that the decrease in CentC abundance seen post-domestication is due to selection for smaller active centromeres to complement the smaller overall genome size.  While our current data are insufficient to evaluate this conclusion, future work investigating differences in CentC content among maize landraces that vary in genome size (CITE) may provide an opportunity to further test this hypothesis.

@@

% For tables use
\begin{table}[]
\caption{ChIP Reads mapping to Unassembled from different Oat-Maize Addition (OMA) Lines.  Reads were mapped using Bowtie2 \citep{langmead2012} with the paramater -very sensitive-local.}\label{supp.gaby}       % Give a unique label

>>>>

\institute{Paul Bilinski : Kevin Distor : Jose Gutierrez-Lopez : Gabriela Mendoza Mendoza : Jeffrey Ross-Ibarra\at
              Department of Plant Sciences, UC Davis\\
              Davis, California, 95616
              \and
              Jeffrey Ross-Ibarra\at
              The Genome Center and Center for Population Biology, University of California\\
              Davis, California, 95616\\
              \email{rossibarra@ucdavis.edu}           %  \\
              \and
              Jose Gutierrez-Lopez\at
              Department of Natural Resources and the Environment
              University of New Hampshire\\
              Durham, New Hampshire, 03820
%             \emph{Present address:} of F. Author  %  if needed
           \and
           Jinghua Shi : R. Kelly Dawe\at
           Department of Plant Biology, University of Georgia\\
	  Athens, Georgia, 30602
           \and
           R. Kelly Dawe\at
           Department of Genetics, University of Georgia\\
           Athens, Georgia, 30602\\
}

@@

Centromere repeats are found in most eukaryotes and play a critical role in kinetochore formation.  Though they exhibit considerable diversity both within and among species, little is understood about the mechanisms that drive centromere repeat evolution.  Here, we use maize as a model to investigate how a complex history involving polyploidy, fractionation, and recent domestication has impacted the diversity of the maize CentC repeat.  We first validate the existence of long tandem arrays of repeats in maize and other taxa in the genus \emph{Zea}.  We find that genetic similarity among CentC copies is highest within these arrays, suggesting that tandem duplications are the primary mechanism for the generation of new copies.  In spite of this, we see little evidence that CentC variants form distinct genetic groups, but instead find that homoplasious mutations may have homogenized CentC diversity.   Although the two ancestral subgenomes of maize have contributed nearly equal numbers of centromeres, our analysis shows that a vast majority of all CentC repeats derive from only one of the parental genomes.  Finally, by comparing maize with its wild progenitor teosinte, we find that the abundance of CentC has decreased through domestication while the pericentromeric repeat Cent4 has drastically increased.

@@

\section*{Introduction}
\label{intro}
In spite of the rapid growth of sequenced plant genomes, plant centromeres remain poorly understood and relatively cryptic due to their highly repetitive content.  Centromere repeats are highly diverse across taxa and their turnover appears to be very rapid \citep{Melters2012}. However, little is known about the genetic mechanisms that produce centromere repeat diversity.  
Domesticated maize (\emph{Zea mays} ssp. \emph{mays}) has a high quality genome assembly \citep{Schnable2009} including complete sequence of two centromeres \citep{Wolfgruber2009}, and the breadth of research into maize centromeres makes it one of the best systems in which to investigate the processes governing centromere repeat evolution.

@@

%We wanted to better understand the diversity present in all CentC copies across the genome.  Looking at our alignment, we found that the sequences had a pairwise percent identify of 65.8\% with only 2.2\% of sites identical, suggesting large sequence diversity.  However, these statistics do not inform us whether the amount of diversity observed would be expected due to cumulative mutation over time.  
In order to better understand patterns of diversity at CentC, we performed simulations to test the likelihood of homoplasious mutations (i.e. independent mutations occurring at the same position in two different CentC repeats). Our simulation (code available at: \url{https://github.com/paulbilinski/CentC_Analyses/tree/master/Diversity_sims}) assumed that CentC has been evolving for  1 million years since the divergence of maize and \emph{Tripsacum} \citep{Ross-Ibarra2009}, a closely related genus whose centromere repeat shares a large amount of homology \citep{Melters2012}.  We assumed a constant copy number, a mutation rate of $3 x 10^{-8}$ \citep{clark2005estimating}, and one generation per year.

@@

\begin{figure}
% For two-column wide figures use
% Use the relevant command to insert your figure file.
%  \includegraphics[width=0.75\textwidth]{circos.png}
\includegraphics{Fig1_circos}
\caption{CentC repeat location in relation to the maize subgenomes.  The outer ring depicts chromosomal assignment to the two subgenomes, with higher confidence regions are colored with darker colors.  Breakpoints between the subgenomes remain uncolored to indicate uncertainty. The middle ring, shaded in blue, displays the locations of all CentCs across the genome.  The inner ring, shaded in yellow, displays the locations of all CentC clusters greater than 20KB in length}
\label{circos}    
\end{figure}

@@

\begin{figure}[h]
\centering
% For two-column wide figures use
% Use the relevant command to insert your figure file.
\includegraphics[width=1\textwidth]{Fig3_Heatmap}
\caption{CentC physical location and genetic relatedness for (a) chromosome 2 and (b) chromosome 5.  On the physical map, red lines show locations of numbered CentC clusters and blue blocks show the location of the active kinetochores.  Scale bar is in MB.  Below each physical map is shown a heatmap of genetic relatedness of each CentC to (top row) other copies within its island of tandem repeats delineated by dotted lines and (bottom row) all other copies on the chromosome.  Darker colors indicate higher relatedness.  The total number of CentC in each cluster is shown below the map
}
\label{heatmap}    
\end{figure}

@@

Our analysis of centromere repeat diversity across the maize genome identifies thousands of copies exhibiting tremendous diversity. 
But while we can cluster the repeats into groups of related sequences, these groups seem to have little relation to current or ancient maize chromosomes (Figure \ref{pcoa}).  
We see no evidence of chromosome specific repeats such as that seen in \emph{Arabidopsis} species \citep{Kawabe2005, Pontes2004}, suggesting the presence of a mechanisms that homogenizes repeats across centromeres on different chromosomes. 
We further show that Cent4, once thought to be a chromosome-specific centromere repeat \citep{Page2001}, appears to be a poorly characterized tandem repeat or nonautonomous retroelement, but is not associated with the centromere.  
Inter-chromosomal transposition was postulated by \citet{Shi2010} as an explanation for marker genotypes showing evidence of recombination, and such transposition could also explain the similarity of repeats across all chromosomes.

Instead of chromosome-specific groups, we find that virtually all the large arrays of CentC in the maize  genome derive from one of the two ancestral genomes present in modern day \emph{Zea} (Figure \ref{circos}).  
One possible explanation for the biased inheritance pattern is that centromere inactivation mirrors the differential expression seen between the subgenomes.  
Centromere inactivation could then lead to differential loss of CentC clusters, such that large clusters remain only in the active centromeres.
Large CentC clusters are clearly biased in favor of subgenome 1, which is known to have lower gene loss and higher average expression than maize subgenome 2 \citep{Schnable2011}. 
Consistent with a correlation between expression and CentC loss in subgenome 2, \citet{ReinhartBartel2002} find evidence of regulatory microRNA’s corresponding to centromeric repeats.   
An alternative explanation is simply that the subgenome 1 parent had more or larger CentC arrays, leading to an initial unequal representation in the ancient polyploid ancestor of maize.

@@

%!RELATEDNESS
Our sequence comparison of CentCs also enabled us to explore the relationship between genetic and physical distance among repeats.  
Using the well-assembled centromeres on chromosomes 2 and 5, we found spatial autocorrelation of relatedness among repeats, and we observe genome-wide that many CentC’s within an array fall into the same genetic cluster.  
Our observations are consistent with the simple idea that most repeats arise due to tandem duplication, and long distance transposition of CentC, while necessary to homogenize repeats across chromosomes \citep{Shi2010}, is relatively uncommon.

@@

Previous cytogenetic work  identified  differences in centromere repeat content  between domesticated maize and its wild relatives \emph{Z. mays} ssp. \emph{parviglumis}, \emph{Z. mays} ssp. \emph{mexicana}, and \emph{Z. luxurians} (hereafter \emph{parviglumis}, \emph{mexicana,} and \emph{luxurians}) \citep{Albert2010} but was unable to quantify  differences. 
Our resequencing results show that while there is little difference in the distribution of CentC in tandem arrays, the absolute abundance of CentC has  decreased during domestication, and we verify this with FISH in two maize-teosinte hybrid individuals (Figure \ref{abundance}). 
Decreasing CentC content during domestication contrasts with the apparent lack of change in transposable element abundance after controlling for genome size \citep{Chia2012}, suggesting that changes in CentC are not due to causes common to all repetitive sequence.
The maize genome is smaller than its teosinte counterpart, largely due to differences in the abundance of heterochromatic knobs \citep{poggio1998}.  \citet{Zhang2012} have postulated an adaptive relationship between centromere size and genome size based on an observed correlation between centromere size and genome size across a number of grass species.  
A model correlating centromere size to total genome size would propose that the decrease in CentC abundance seen post-domestication is due to selection for smaller active centromeres to complement the smaller overall genome size.
While our current data are insufficient to evaluate this conclusion, future work investigating differences in CentC content among maize landraces that vary in genome size \citep{poggio1998} may provide an opportunity to further test our hypothesis.

@@

% For tables use
\begin{table}[]
\caption{ChIP Reads mapping to Unassembled from different Oat-Maize Addition (OMA) Lines.  Reads were mapped using Bowtie2 \citep{langmead2012fast} with the paramater -very sensitive-local.}\label{supp.gaby}       % Give a unique label
